Lessons learned from bugs in models of human history

### 2020-05-20 paper.tex
small edits and added comment on graphical inspectin

%\sgcomment{How to avoid simulating the wrong demographic model}

\abstract{
Simulation plays a central role in population genomics studies. Recent years
have seen rapid improvements in software efficiency that make it possible to simulate
large genomic regions for many individuals sampled from large numbers of populations.
As the complexity of demographic models we study grows, there is an ever-increasing
opportunity to introduce bugs in their implementation.
Here we describe two errors made in defining population genetic models using
the \msprime\ coalescent simulator that have found their way into the published record.
We discuss how these errors have affected downstream analyses and give recommendations
for software developers and users to reduce the risk of such errors.
}

@@

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used in genetics studies.
Much of its appeal is the large increase in efficiency over the classical
\ms\ program~\citep{hudson2002generating} which makes it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is the Python API that is its primary interface, greatly
increasing the flexibility and ease of use over the standard approach of
text-based command line interfaces.
In particular, programs like \ms\ required users to specify cryptic command
line options to describe demographic models.
Particularly for the large models we use today, these are not
intuitive or comprehensible for humans.
The Python API for \msprime\ is a great improvement, allowing the user to
state models in a documented and reproducible manner.
%\sgcomment{Other software, such as SLiM, 
%have developed advanced user interfaces and even dedicated programming languages 
%to facilitate the specification of various modes of evolution.\cite{}
%}

Implementing multi-population models of demographic history is still hard and error
prone. In this note we discuss two implementation errors that arose through
unfortunate design decisions in \msprime's demography API and which then found their way
into the scientific record. The first error has relatively mild effects on genetic diversity but was 
used in many publications, while the second error was used only once but had a large impact on the 
simulation results.  
In light of these implementation errors, we discuss improvements to \msprime's API motivated by these
discoveries and more general best practices for implementing and simulating complex multi-population
demography.

@@

To illustrate the demography API, \msprime\ included a description of a widely-used
three population Out-of-Africa model~\citep{gutenkunst2009inferring}
as part of its tutorial documentation. In this model (Fig.~\ref{fig:ooa_stats}A),
Eurasian (CEU and CHB) and African (YRI) populations split from each other in the deep past,
followed by a more recent split of European and Asian populations, with variable rates of
continuous migration between each of the populations. However, the implementation in the
\msprime\ tutorial was incorrect. Before the time of the split of African and Eurasian
populations, when there should have been just a single randomly mating population, migration was
allowed to occur between the ancestral population and a second population with size equal to
the Eurasian bottleneck size for all time into the past
(Fig.~\ref{fig:ooa_stats}B). This incorrect model was introduced into
the tutorial for \msprime\~ver.~0.3.0 and remained in the documentation for
around four years.

Fortunately, the effects of this error are subtle.
Population sizes and structure since the time of
the split are unaffected, so differences in expected $F_{ST}$ are negligible between
the correct and incorrect model. However, the ancient structure distorts the distribution
of early TMRCAs. The extraneous ancient population increases the long-term effective 
population size, resulting in roughly $4\%$ excess heterozygosity in 
contemporary populations compared to the correctly specified model
(Fig.~\ref{fig:ooa_stats}C-F).

It is fortunate that the error in the model description has quite a
small effect because the tutorial code has been copied many times
and used in publications. By searching for some identifying strings
from the model definition on GitHub,
we found 32 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repositories to alert the authors.)
In most cases the publications used simulations to test some inference
method and were not directly concerned with the details of the demography,
and the model was simply used as an example of a complex population
history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
\cite{zhou2018popdemog} used the incorrect model as an example
of how their method for visualising demographic models can support
\msprime\ input.
% TODO ask Franz about this
Finally, \cite{pfaffelhuber2020choose}
used simulations of the incorrect model demography to evaluate
their method for choosing ancestry informative markers.
Given the very subtle effect of the incorrect
model on demography (and the fact the method was evaluated using other
simulations and real data), it seems unlikely that the model details
had any qualitative effect on their conclusions.

@@

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
After these initial checks were made, however, the simulation
was performed without passing the parameter of demographic events,
so that the three populations
never merged as expected and remained separated with low levels of migration (Fig~\ref{fig:prs}A),
leading to a vast overestimate of the divergence across human populations.
While the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: heterozygosity was more than
doubled in African populations but just $30-40\%$ the expected
levels in Eurasian populations when compared to diversity under the correctly specified model.

@@

Difficulties in transferability remain in the corrected model (Fig.~\ref{fig:prs}), although
risk prediction in each population is significantly improved when using the correct
demographic model (compare to Fig.~5 in~\citet{martin2017human}).
The corrected simulations indicate that the accuracy of genetic risk scores is still
substantially reduced in understudied populations (Fig.~\ref{fig:prs}E-G),
supporting the main conclusion of~\citet{martin2017human}.
However, the reduction is much less pronounced than reported.
In particular, we do not observe large differences in mean risk prediction
across populations (Fig.~\ref{fig:prs}C,D), as presented in~\citet{martin2017human}.
At least for the neutral polygenic architectures considered here, genetic drift alone
does not appear to induce large directional biases in mean predicted risk across ethnicities.

@@

>>>>

\sgcomment{How to avoid simulating the wrong demographic model}

\abstract{
Simulation plays a central role in population genomics studies. Recent years
have seen rapid improvements in software efficiency that make it possible to simulate
large genomic regions for many individuals sampled from large numbers of populations.
As the complexity of demographic models we study grows, however, there is an ever-increasing
opportunity to introduce bugs in their implementation.
Here we describe two errors made in defining population genetic models using
the \msprime\ coalescent simulator that have found their way into the published record.
We discuss how these errors have affected downstream analyses and give recommendations
for software developers and users to reduce the risk of such errors.
}

@@

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used in genetics studies.
Much of its appeal is the large increase in efficiency over the classical
\ms\ program~\citep{hudson2002generating} which makes it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is the Python API that is its primary interface, greatly
increasing the flexibility and ease of use over the standard approach of
text-based command line interfaces.
In particular, programs like \ms\ required users to specify cryptic command
line options to describe demographic models.
Particularly for the large models we use today, these are not
intuitive or comprehensible for humans.
The Python API for \msprime\, by contrast, allows the user to
state models in a more human-readable manner: 
Example of model specification for the two softwares is provided in the Methods section. 
%\sgcomment{Other software, such as SLiM, 
%have developed advanced user interfaces and even dedicated programming languages 
%to facilitate the specification of various modes of evolution.\cite{}
%}

Implementing multi-population models of demographic history for either software is hard and error-prone. 
In this note we discuss two implementation errors that arose through
unfortunate design decisions in \msprime's demography API and which then found their way
into the scientific record. The first error has relatively mild effects on genetic diversity but was 
used in many publications, while the second error was used only once but had a large impact on the 
simulation results.  
In light of these implementation errors, we discuss improvements to \msprime's API motivated by these
discoveries and, more generally, best practices for implementing and simulating complex multi-population
demography.

@@

To illustrate the demography API, \msprime\ included a description of a widely-used
three population Out-of-Africa model~\citep{gutenkunst2009inferring}
as part of its tutorial documentation. In this model (Fig.~\ref{fig:ooa_stats}A),
Eurasian (CEU and CHB) and African (YRI) populations split from each other in the deep past,
followed by a more recent split of European and Asian populations, with variable rates of
continuous migration between each of the populations. However, the implementation in the
\msprime\ tutorial was incorrect. Before the time of the split of African and Eurasian
populations, when there should have been just a single randomly mating population, migration was
allowed to occur between the ancestral population and a second population with size equal to
the Eurasian bottleneck size for all time into the past
(Fig.~\ref{fig:ooa_stats}B). This incorrect model was introduced into
the tutorial for \msprime~ver.~0.3.0 and remained in the documentation for
around four years.

Fortunately, the effects of this error are subtle.
Population sizes and structure since the time of
the earliest split are unaffected, so differences in expected $F_{ST}$ are negligible between
the intended and incorrect models. However, the ancient structure distorts the distribution
of early TMRCAs. The extraneous ancient population increases the long-term effective 
population size, resulting in roughly $4\%$ excess heterozygosity in 
contemporary populations compared to the intended model
(Fig.~\ref{fig:ooa_stats}C-F).

Even though the error had a limited effect on simulated data, the tutorial code has been 
copied many times and used in publications. By searching for some identifying strings
from the model definition on GitHub,
we found 32 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repositories to alert the authors.)
In most cases the publications used simulations to test some inference
method and were not directly concerned with the details of the demography,
and the model was simply used as an example of a complex population
history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
\cite{zhou2018popdemog} used the incorrect model as an example
of how their method for visualising demographic models can support
\msprime\ input.
% TODO ask Franz about this
Finally, \cite{pfaffelhuber2020choose}
used simulations of the incorrect model demography to evaluate
their method for choosing ancestry informative markers.
Given the very subtle effect of the incorrect
model on demography (and the fact the method was evaluated using other
simulations and real data), it seems unlikely that the model details
had any qualitative effect on their conclusions.

@@

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
After these initial checks were made, however, the simulation
was performed without passing the parameter of demographic events,
so that the three populations
never merged as expected and remained separated with low levels of migration (Fig~\ref{fig:prs}A),
leading to a vast overestimate of the divergence across human populations.
While the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: heterozygosity was more than
doubled in African populations and reduced by more than half in Eurasian populations 
relative to the intended model.

@@

Difficulties in transferability remain in the intended model (Fig.~\ref{fig:prs}), although
risk prediction in each population is significantly improved (compare to Fig.~5 in~\citet{martin2017human}).
The intended simulations indicate that the accuracy of genetic risk scores is still
substantially reduced in understudied populations (Fig.~\ref{fig:prs}E-G),
supporting the main conclusion of~\citet{martin2017human}.
However, the reduction is much less pronounced than reported.
In particular, we do not observe large differences in mean risk prediction
across populations (Fig.~\ref{fig:prs}C,D), as presented in~\citet{martin2017human}.
At least for the neutral polygenic architectures considered here, genetic drift alone
does not appear to induce large directional biases in mean predicted risk across ethnicities.

@@

Graphical inspection of demographic models can also help identify issues, especially if 
visualization can be automated. msprime does contain a demography debugging tool summarizing 
demographic events occurring over time, and various efforts are underway to facilitate 
the specification and visualization of population genetics models \cite{zhou2018popdemog}, \sgcomment{Mention Aaron's code and Slim?}.
